Summary Statistics

### 2025-03-01 extensive_tests/draft-sumstat-vignette.Rmd
Updating functions for fitch estimation per island and corresponding tests

### Frequency of Methylation State Changes at Single Sites

We compare the methylation state of single sites at the tree cherries. A cherry is a pair of leaf nodes (also called tips or terminal nodes) in a phylogenetic tree that share a direct common ancestor. In other words, if two leaves are connected to the same internal node and no other leaves are connected to that internal node, they form a cherry.

We calculate the frequency of sites that changes for each structure (island and non-island structures).

Let $S_k$ be the set of sites in the cherry structure of interest, where $k$ represents the index of the cherry structure. For each site in $S_k$, let the methylation state at a given tip $t$ be denoted as $m_{k,s,t}$, where $s \in S_k$ is the index of the site, and $m_{k,s,t} \in {0, 0.5, 1}$ represents the methylation state at tip $t$ (unmethylated, partially-methylated, or methylated, respectively).

For a given cherry structure, we compute in the proportion of sites that have a different methylation state between two tips, say $t_1$ and $t_2$.

Define the indicator variable:
\[
\delta_{k,s,t_1,t_2} =
\begin{cases}
1, & \text{if } m_{k,s,t_1} \neq m_{k,s,t_2}, \\
0, & \text{if } m_{k,s,t_1} = m_{k,s,t_2}.
\end{cases}
\]

Then, the summary statistic for the proportion of sites with different methylation states between tips $t_1$ and $t_2$ in the cherry structure $k$ is:
\[
\text{Proportion}_k(t_1, t_2) = \frac{1}{|S_k|} \sum_{s \in S_k} \delta_{k,s,t_1,t_2}
\]

where $|S_k|$ is the total number of sites in the cherry structure $k$, and the sum counts the number of sites with different methylation states between tips $t_1$ and $t_2$.

This statistic gives the proportion of sites in the cherry structure $k$ where the methylation state differs between the two tips of interest.

@@

>>>>

### Mean Frequency of Methylation State Changes at Single Sites per Structure Type

Let \(S_k\) be the set of sites in the genomic structure indexed by \(k\). For each site \(s \in S_k\), let the methylation state at a given tip \(t\) be denoted as \(m_{k,s,t}\), where \(m_{k,s,t} \in \{0, 0.5, 1\}\) represents the unmethylated, partially-methylated, or methylated state, respectively.

For a given cherry \((t_1, t_2)\), we compute the proportion of sites with differing methylation states between the two tips.

Define the indicator variable:
$$
\delta_{k,s,t_1,t_2} =
\begin{cases}
1, & \text{if } m_{k,s,t_1} \neq m_{k,s,t_2}, \\
0, & \text{if } m_{k,s,t_1} = m_{k,s,t_2}.
\end{cases}
$$

Then, the proportion of sites with different methylation states in structure \(k\) for the cherry \((t_1, t_2)\) is given by:

$$
F_k(t_1, t_2) = \frac{1}{|S_k|} \sum_{s \in S_k} \delta_{k,s,t_1,t_2}
$$

where \(|S_k|\) is the total number of sites in structure \(k\), and the sum counts the number of sites where the methylation state differs between tips \(t_1\) and \(t_2\).

This statistic quantifies the proportion of differing methylation states for a given genomic structure in a cherry.

To compute the weighted mean frequency of methylation changes across all structures in a cherry, we separate island and non-island structures.

For a given cherry \((t_1, t_2)\), let \( \mathcal{I} \) be the set of indices for island structures and \( \mathcal{N} \) be the set of indices for non-island structures. Each structure \( k \) has a total number of sites \( |S_k| \), which we use as weights.

The weighted mean frequency of methylation changes for island structures is:

$$
\bar{F}_{\text{island}}(t_1, t_2) = \frac{\sum_{k \in \mathcal{I}} |S_k| F_k(t_1, t_2)}{\sum_{k \in \mathcal{I}} |S_k|}
$$

Similarly, the weighted mean frequency for non-island structures is:

$$
\bar{F}_{\text{non-island}}(t_1, t_2) = \frac{\sum_{k \in \mathcal{N}} |S_k| F_k(t_1, t_2)}{\sum_{k \in \mathcal{N}} |S_k|}
$$

These expressions compute the mean frequency of methylation state differences across all island and non-island structures, weighted by the number of sites in each structure.

@@

# Fitch

- `get_meanMeth_islands` computes the mean methylation state of all the islands. **Maybe this should allow excluding island borders**
